---
title: "Introduction to BioHEART Resilience Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BioHEART Resilience Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `BioHEARTResilience` package implements a novel methodology to identify cardiovascular disease resilience phenotypes by comparing observed coronary artery calcium scores (CACS) against expected values derived from ensemble cardiovascular risk scoring.

This approach, developed from the BioHEART study, allows researchers to classify patients as:

- **Resilient**: Lower CACS than expected given their risk profile
- **Susceptible**: Higher CACS than expected given their risk profile  
- **Reference**: CACS close to what's expected
- **Other**: Intermediate classifications

## Key Features

- **Multiple Risk Scores**: Integrates FRS, ASCVD, MESA, and SCORE2 risk calculations
- **Ensemble Methodology**: Combines risk scores using ordered quantile normalization
- **Zero-Inflated Modeling**: Accounts for the high prevalence of zero CACS values
- **Comprehensive Visualization**: Publication-ready plots and diagnostics
- **Flexible Analysis**: Handles missing data and customizable thresholds

## Installation

```{r eval=FALSE}
# Install from GitHub (when available)
devtools::install_github("yourusername/BioHEARTResilience")

# Load the package
library(BioHEARTResilience)
```

```{r setup}
library(BioHEARTResilience)
```

## Quick Start: Complete Analysis

The easiest way to perform a resilience analysis is using the main `resilience_analysis()` function:

```{r}
# Load example data
data("example_cohort")

# Examine the data structure
str(example_cohort)
```

Let's look at the first few rows:

```{r}
head(example_cohort)
```

And check the data summary:

```{r}
summary(example_cohort[c("age", "tc", "hdl", "sbp", "cacs")])
```

Now let's run the complete resilience analysis:

```{r eval=FALSE}
# Run complete analysis
results <- resilience_analysis(
  data = example_cohort,
  cacs_col = "cacs",
  cholesterol_unit = "mmol/L",
  include_plots = TRUE
)
```

```{r echo=FALSE}
# For vignette, run without plots to avoid potential issues
results <- resilience_analysis(
  data = example_cohort,
  cacs_col = "cacs", 
  cholesterol_unit = "mmol/L",
  include_plots = FALSE
)
```

The function provides detailed progress reporting:

```{r}
# View the results summary
print(results)
```

## Accessing Results

The results object contains all analysis components:

### Classification Results

```{r}
# View resilience classifications
table(results$classifications$resilience_class)
```

```{r}
# Look at classification details for first 10 subjects
head(results$classifications[c("cacs_original", "cacs_percentile", "resilience_class")], 10)
```

### Risk Scores

```{r}
# Check risk score completion
head(results$risk_summary$summary)
```

### Model Performance

```{r}
# Model diagnostics
results$model_diagnostics$fit_statistics
```

## Step-by-Step Analysis

For more control, you can run each step individually:

### Step 1: Data Preparation

```{r}
# Prepare data with validation
prepared_data <- prepare_cohort_data(
  example_cohort,
  cholesterol_unit = "mmol/L"
)

cat("Prepared data dimensions:", dim(prepared_data), "\n")
```

### Step 2: Calculate Risk Scores

```{r}
# Calculate individual risk scores
risk_scores <- calculate_risk_scores(
  prepared_data,
  scores = c("frs", "ascvd", "mesa", "score2")
)

# Summarize risk scores
risk_summary <- summarize_risk_scores(risk_scores)
print(risk_summary)
```

### Step 3: Create Ensemble Score

```{r}
# Create ensemble risk score using ordered quantile normalization
ensemble_scores <- ensemble_risk_score(risk_scores)

# Check ensemble results
cat("Ensemble scores created for", sum(!is.na(ensemble_scores$average_norm_score)), "subjects\n")
head(ensemble_scores[c("average_norm_score", "n_scores_used")])
```

### Step 4: Fit CACS Model

```{r}
# Combine data for modeling
modeling_data <- merge(prepared_data, ensemble_scores, by = "row.names")
rownames(modeling_data) <- modeling_data$Row.names
modeling_data$Row.names <- NULL

# Fit zero-inflated negative binomial model
cacs_model <- fit_cacs_model(
  modeling_data,
  cacs_col = "cacs",
  risk_score_col = "average_norm_score"
)

# Check model summary
summary(cacs_model)
```

### Step 5: Calculate Risk-Adjusted Percentiles

```{r}
# Calculate the "Calcium Vulnerability Score"
percentiles <- calculate_cacs_percentiles(
  cacs_model,
  modeling_data
)

# Examine percentile distribution
summary(percentiles$cacs_percentile)
```

### Step 6: Classify Resilience

```{r}
# Classify based on percentile thresholds
classifications <- classify_resilience(percentiles)

# View classification distribution
table(classifications$resilience_class)
```

## Visualization

The package provides several visualization functions:

### CACS vs Risk Score Plot

```{r eval=FALSE}
# Combine final data
final_data <- merge(modeling_data, classifications, by = "row.names")
rownames(final_data) <- final_data$Row.names

# Plot CACS vs risk score colored by classification
plot_cacs_vs_risk(
  final_data,
  cacs_col = "cacs",
  risk_score_col = "average_norm_score", 
  class_col = "resilience_class"
)
```

### Risk Score Distribution

```{r eval=FALSE}
# Plot risk score distributions by class
plot_risk_distribution(
  final_data,
  risk_score_col = "average_norm_score",
  class_col = "resilience_class"
)
```

### Model Diagnostics

```{r eval=FALSE}
# Generate model diagnostic plots
diagnostic_plots <- plot_model_diagnostics(cacs_model, percentiles)
```

## Customizing the Analysis

### Custom Risk Scores

```{r eval=FALSE}
# Use only specific risk scores
results_custom <- resilience_analysis(
  example_cohort,
  risk_scores = c("frs", "ascvd"),  # Only FRS and ASCVD
  include_plots = FALSE
)
```

### Custom Classification Thresholds

```{r eval=FALSE}
# Use custom percentile thresholds
results_custom_thresh <- resilience_analysis(
  example_cohort,
  percentile_thresholds = c(
    resilient = 15,        # More stringent resilient threshold
    reference_low = 35,
    reference_high = 65, 
    susceptible = 85       # More stringent susceptible threshold
  ),
  include_plots = FALSE
)
```

### Different Risk Regions (for SCORE2)

```{r eval=FALSE}
# For European populations at moderate risk
results_moderate_risk <- resilience_analysis(
  example_cohort,
  risk_region = "Moderate",  # For SCORE2 calculation
  include_plots = FALSE
)
```

## Handling Missing Data

The package gracefully handles missing data with informative warnings:

```{r}
# Check missing data patterns in original data
missing_summary <- check_missing_values(example_cohort, verbose = FALSE)
if (nrow(missing_summary) > 0) {
  print(missing_summary)
} else {
  cat("No missing data found in example dataset\n")
}
```

## Advanced Usage

### Getting Risk Score Information

```{r}
# Get information about available risk scores
risk_score_info()
```

### Validating Data for Risk Scores

```{r}
# Check which subjects are valid for each risk score
valid_subjects <- validate_risk_data(
  prepared_data, 
  scores = c("frs", "ascvd", "mesa", "score2")
)

cat("Valid subjects for all scores:", sum(valid_subjects), "out of", length(valid_subjects), "\n")
```

### Individual Risk Score Calculation

```{r eval=FALSE}
# Calculate individual risk scores manually
frs_scores <- CVrisk::ascvd_10y_frs(
  gender = prepared_data$gender,
  age = prepared_data$age,
  hdl = prepared_data$hdl_mgdl,
  totchol = prepared_data$tc_mgdl,
  sbp = prepared_data$sbp,
  bp_med = prepared_data$bp_med,
  smoker = prepared_data$curr_smok,
  diabetes = prepared_data$cvhx_dm
)
```

## Interpreting Results

### Resilience Classifications

- **Resilient (< 20th percentile)**: Subjects with unexpectedly low CACS given their cardiovascular risk profile. These individuals may have protective factors worth investigating.

- **Susceptible (> 80th percentile)**: Subjects with unexpectedly high CACS given their risk profile. These individuals may have additional risk factors or genetic susceptibility.

- **Reference (40-60th percentile)**: Subjects whose CACS aligns with expectations based on traditional risk factors.

- **Other**: Subjects in intermediate percentile ranges.

### Clinical Implications

Resilient and susceptible phenotypes represent extremes of the risk-to-disease relationship and may harbor novel protective or risk factors that could inform:

- Biomarker discovery
- Therapeutic target identification  
- Precision medicine approaches
- Understanding of disease mechanisms

## Best Practices

1. **Data Quality**: Ensure high-quality CACS measurements and clinical data
2. **Missing Data**: Review missing data patterns and consider imputation if systematic
3. **Model Validation**: Check model diagnostics, especially percentile uniformity
4. **Threshold Selection**: Consider population-specific thresholds based on study goals
5. **Replication**: Validate findings in independent cohorts when possible

## Getting Help

For questions about the package:

```{r eval=FALSE}
# View function help
?resilience_analysis
?calculate_risk_scores
?ensemble_risk_score

# View this vignette
vignette("introduction", package = "BioHEARTResilience")
```

## References

1. BioHEART Study: [Citation information will be added]
2. CVrisk package: Castro V et al. CVrisk: Compute Risk Scores for Cardiovascular Diseases. R package.
3. RiskScorescvd package: Risk score calculations for cardiovascular disease.
4. Zero-inflated models: Zeileis A, Kleiber C, Jackman S. Regression Models for Count Data in R. Journal of Statistical Software.

---

*This vignette demonstrates the basic usage of the BioHEARTResilience package. For more advanced applications and detailed methodology, please refer to the original research publication.*